Bayesian Deep Gaussian Processes for Correlated Functional Data: \\ A Case Study in Cosmological Power Spectra

### 2025-05-23 paper/main.tex
annie done through 3.2

Let $S$ represent the true (but unknown) underlying power matter spectra for
a particular cosmology.  We consider $\bar{Y}$ a noisy realization of $S$.
Specifically, we assume $\bar{Y}$ is a random realization of a Gaussian 
process centered at $S$ with some covariance, i.e. 
$\bar{Y}|S \sim \mathrm{GP}(S, \Sigma_\varepsilon)$.  While previous works
have considered diagonal $\Sigma_\varepsilon$ \citep{moran2024dpc}, we find it
essential to account for the smoothly varying spatial dependence of the
low resolution runs (as visualized in Figure \ref{fig:plot_data}).  For
this, we embrace variations of the Mat\'ern kernel \citep{stein1999interpolation},
tailored to each scenario (more on this in later subsections).

We next place a zero-mean Gaussian process prior on $S$, i.e.,
$S\sim \mathrm{GP}\left(\mathbf{0}, \Sigma_S(\cdot)\right)$, again with a 
spatial kernel.  
\textcolor{red}{(@Steve - is a zero mean the right thing?  If not, what is the mean?)}
Standard practice would be to use locations $X$ as the inputs to kernel $\Sigma_S(\cdot)$.
Yet we have found that prior to lack sufficient flexibility for our motivating
application.  In the Mira-Titan setting, some nonstationarity is expected due to 
baryonic acoustic oscillations when $-2 \leq \log_{10}(k) \leq -1$. To account 
for this, we incorporate an additional latent Gaussian process, $W$, which accommodates
nonstationarity by warping $X$.  Ultimately, this yields the following hierarchical model:
\begin{equation}\label{eq:dgphm}
\begin{aligned}
\bar{Y}|S &\sim \mathrm{GP}(S,\Sigma_\varepsilon) \\
S|W &\sim \mathrm{GP}\left(\mathbf{0}, \Sigma_S(W)\right) \\
W &\sim \mathrm{GP}\left(\mathbf{0}, \Sigma_W(X)\right)
\end{aligned}
\end{equation}
$W$ is also an $n$-vector whose entries $w_i$ correspond to warped versions of each $x_i$.
The functional composition of GPs in the latter two lines forms a ``deep Gaussian process'' 
\citep{damianou2013deep,dunlop2018deep}. Additional latent layers could be considered, 
although we find one is sufficient.

We use the Mat\`ern kernels,
\[
\begin{aligned}
&\Sigma_S(W)^{i,j} = \tau^2 \left( 1 + \frac{\sqrt{5}d}{\sqrt{\theta_S}} + 
  \frac{5d^2}{3\theta_S}\right) \exp\left(-\frac{\sqrt{5}d}{\sqrt{\theta_S}}\right)
\quad\textrm{for}\quad 
d=||w_i-w_j||^2, \;\;\textrm{and} \\
&\Sigma_W(X)^{i,j} = \left( 1 + \frac{\sqrt{5}d}{\sqrt{\theta_W}} + 
  \frac{5d^2}{3\theta_W}\right) \exp\left(-\frac{\sqrt{5}d}{\sqrt{\theta_W}}\right)
\quad\textrm{for}\quad 
d=||x_i-x_j||^2.
\end{aligned}
\]
On the $S$ layer, hyperparameters $\tau^2$ and $\theta_S$ govern the 
correlation scale and lengthscale respectively.  On the $W$ layer, we allow for a
unique lengthscale $\theta_W$ but enforce unit scale following \citet{sauer2023active}.
We considered setting the prior mean of $W$ to $X$ instead of $\mathbf{0}$
\citep[which would indicate stationarity apriori;][]{schmidt2003bayesian}, but found
it had a neglible effect on the model fits.

@@

We transform $X$ values to fall within $[0,1]$ and $\bar Y$ so that the mean and variance 
are 0 and 1, respectively. This allows for a straightforward specification of priors, 
where the hyperparameters for the Mat\'ern covariance $\Sigma_S$ each take a vague prior. 
Specifically:

where. From here, we estimate $\tau_\ell, \theta_\ell$ with maximum 
likelihood, and conditional on $\Sigma_\varepsilon$, we specify priors for $\tau_S, \theta_S$ 
following \cite{sauer2023active}: \textcolor{red}{Annie, please ensure the most up-to-date 
prior specs are shown above and here. Here is what I found from fit\$settings when using 
deepgp v1.1.3: $\theta_S \sim G(\alpha=1.2, \beta=1)$ and $\tau_S \propto 1/\tau_S^2$.}

A key property of GPs is that any finite set of $n$ observations from the GP is a 
multivariate normal distribution. Therefore, upon integrating out $S$, we obtain 
$\bar{Y} \sim \mathcal{N}_n(\mu, \Sigma_S+\Sigma_\varepsilon)$ for the $n$ observed X locations. 
Additionally, this model yields the posterior distribution $S|\bar Y \sim \mathcal{N}_n(m, C)$, 
where $C=\left(\Sigma_S^{-1}+\Sigma_\varepsilon^{-1}\right)^{-1}$ and 
$m=C\left(\Sigma_\varepsilon^{-1}\bar Y+\Sigma_S^{-1}\mu\right)$. 
This distribution provides the posterior mean and corresponding uncertainty for 
$S$ within a given cosmology. 
% With this hierarchical model, we improve the efficiency of estimating $S|\bar{Y}$.

\textcolor{red}{Annie: want to add something here? For $\theta_w$, we also specify 
a Gamma prior of the form $\theta_w \sim G(\alpha=1.2, \beta=2)$. We discuss the 
elliptical slice sampling of $W$, but otherwise it is plugged in to the methods 
of the previous section with very few changes.}

Through the synthesis of the DGP model alongside the additional level of hierarchical 
modeling (HM) to incorporate multiple functional model runs, we call this method DGP.HM.

@@

>>>>

Let $S$ represent the true (but unknown) underlying power matter spectra for
a particular cosmology.  We consider $\bar{Y}$ a noisy realization of $S$.
Specifically, we assume $\bar{Y}$ is a random realization of a Gaussian 
process centered at $S$ with some covariance, i.e. 
$\bar{Y}\mid S \sim \mathrm{GP}(S, \Sigma_\varepsilon)$.  While previous works
have considered diagonal $\Sigma_\varepsilon$ \citep{moran2024dpc}, we find it
essential to account for the smoothly varying spatial dependence of the
low resolution runs (as visualized in Figure \ref{fig:plot_data}).  For
this, we embrace variations of the Mat\'ern kernel \citep{stein1999interpolation},
tailored to each scenario (more on this in later subsections).

We next place a zero-mean Gaussian process prior on $S$, i.e.,
$S\sim \mathrm{GP}\left(\mathbf{0}, \Sigma_S(\cdot)\right)$, again with a 
spatial kernel.  
\textcolor{red}{(@Steve - is a zero mean the right thing?  If not, what is the mean?)}
Standard practice would be to use locations $X$ as the inputs to kernel $\Sigma_S(\cdot)$.
Yet we have found that prior to lack sufficient flexibility for our motivating
application.  In the Mira-Titan setting, some nonstationarity is expected due to 
baryonic acoustic oscillations when $-2 \leq \log_{10}(k) \leq -1$. To account 
for this, we incorporate an additional latent Gaussian process, $W$, which accommodates
nonstationarity by warping $X$.  Ultimately, this yields the following hierarchical model:
\begin{equation}\label{eq:dgphm}
\begin{aligned}
\bar{Y}\mid S,\Sigma_\varepsilon &\sim \mathrm{GP}(S,\Sigma_\varepsilon) \\
S\mid W &\sim \mathrm{GP}\left(\mathbf{0}, \Sigma_S(W)\right) \\
W &\sim \mathrm{GP}\left(\mathbf{0}, \Sigma_W(X)\right)
\end{aligned}
\end{equation}
$W$ is also an $n$-vector whose entries $w_i$ correspond to warped versions of each $x_i$.
The functional composition of GPs in the latter two lines forms a ``deep Gaussian process'' 
\citep{damianou2013deep,dunlop2018deep}. Additional latent layers could be considered, 
although we find one is sufficient.

We use the Mat\`ern kernels,
\[
\begin{aligned}
&\Sigma_S(W)^{i,j} = \tau^2 \left( 1 + \frac{\sqrt{5}d}{\sqrt{\theta_S}} + 
  \frac{5d^2}{3\theta_S}\right) \exp\left(-\frac{\sqrt{5}d}{\sqrt{\theta_S}}\right)
\quad\textrm{for}\quad 
d=||w_i-w_j||^2, \;\;\textrm{and} \\
&\Sigma_W(X)^{i,j} = \left( 1 + \frac{\sqrt{5}d}{\sqrt{\theta_W}} + 
  \frac{5d^2}{3\theta_W}\right) \exp\left(-\frac{\sqrt{5}d}{\sqrt{\theta_W}}\right)
\quad\textrm{for}\quad 
d=||x_i-x_j||^2.
\end{aligned}
\]
In the $S$ layer, hyperparameters $\tau^2$ and $\theta_S$ govern the 
correlation scale and lengthscale respectively.  In the $W$ layer, we allow for a
unique lengthscale $\theta_W$ but enforce unit scale following \citet{sauer2023active}.
We considered setting the prior mean of $W$ to $X$ instead of $\mathbf{0}$
\citep[which would indicate stationarity apriori;][]{schmidt2003bayesian}, but found
it had a neglible effect on the model fits.

@@

While our inferential goal is to obtain the posterior distribution of 
$S\mid\bar{Y},\Sigma_\varepsilon$, this requires inferring latent quantity $W$.  
We first integrate over $S$ to condense our hierarchical model of Eq.~\ref{eq:dgphm} into
\begin{align}
\label{eq:likelihood}
\bar{Y} \mid W, \Sigma_\varepsilon &\sim \textrm{GP}(\mathbf{0}, \Sigma_S(W) + \Sigma_\varepsilon) \\
\label{eq:prior}
W &\sim \mathrm{GP}\left(\mathbf{0}, \Sigma_W(X)\right)
\end{align}
Details are provided in Appendix \textcolor{red}{???}. We then use elliptical slice
sampling \citep[ESS;][]{murray2010elliptical} to generate posterior samples of $W$.  
Proposals are generated from prior samples from Eq.~(\ref{eq:prior}), and acceptances 
involve the ratio of likelihoods from Eq.~\ref{eq:likelihood}.

We also embrace a fully-Bayesian treatment of unknown kernel hyperparameters 
$\{\tau^2, \theta_S, \theta_W\}$.  We prescale $X$ to $[0, 1]$
and $\bar{Y}$ to zero mean and unit variance.  On this scale, we adopt the default 
prior specifications provided in the {\tt deepgp} {\sf R}-package \citep{deepgp}.
\textcolor{red}{(@Steve - check the version number of the package in this reference.  Also,
I'm not sure what we really do with $\tau^2$.  In my package, we assign it a reference prior and
integrate it out.  I don't think we can do that here since we have the $\Sigma_\varepsilon$.)}
%\textcolor{red}{Annie, please ensure the most up-to-date 
%prior specs are shown above and here. Here is what I found from fit\$settings when using 
%deepgp v1.1.3: $\theta_S \sim G(\alpha=1.2, \beta=1)$ and $\tau_S \propto 1/\tau_S^2$.}
We integrate Metropolis-Hastings sampling of these parameters in a Gibbs framework with ESS
sampling of $W$ following the scheme of \citep{sauer2023active}.

Given burned-in posterior samples $\{W^{(t)}, \tau^{2(t)}, \theta_S^{(t)}, \theta_W^{(t)}\}$
for $t\in\mathcal{T}$, we may obtain the
posterior distribution of the true power matter spectra for a particular cosmology as
\[
S^{(t)}\mid\bar{Y}, \Sigma_\varepsilon \sim \mathcal{N}_n(m^{(t)}, C^{(t)})
\quad\textrm{where}\quad
\begin{array}{rl}
C^{(t)}&=\left(\Sigma_S(W^{(t)})^{-1}+\Sigma_\varepsilon^{-1}\right)^{-1} \\
m^{(t)}&=C^{(t)}\left(\Sigma_\varepsilon^{-1}\bar{Y}+\Sigma_S(W^{(t)})^{-1}\mu\right).
\end{array}
\] 
\textcolor{red}{(@Steve - provide details of this derivation.  Also what is $\mu$?  Are we 
going with zero here?  Add details of how we combine these things across all the $t$ samples.)}
We denote our inference procedure for the hierarchical model of Eq.~\ref{eq:dgphm} as
``DGP.HM'' given its deep Gaussian process foundation with additional hierarchical modeling (HM)
structure to incorporate multiple functional outputs.
\textcolor{red}{(Is it too late to change this?  Maybe DGP-FCO for functional correlated output?)}

@@

From here, we estimate $\tau_\ell, \theta_\ell$ with maximum 
likelihood, and conditional on $\Sigma_\varepsilon$,
